NOPT048 Notes

### 2019-06-15 misc/study-notes/nopt048notes.tex
Added graph theory notes and TeX macros

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amsthm,amssymb,amsfonts,tikz,tikzsymbols}
\usepackage[shortlabels]{enumitem}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}

\newcommand{\M}{\mathbb{M}}

\newcommand{\F}{\mathcal{F}}
\newcommand{\pp}{\mathcal{P}}

\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\cone}{cone}
\DeclareMathOperator{\aff}{aff}

\renewcommand\qedsymbol{$\Smiley$}

\newenvironment{problem}[2][Problem]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
\newenvironment{definition}[2][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem*{theorem*}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem*{lemma*}{Lemma}

\begin{document}

\title{NOPT048 Notes}
\author{Marcel Goh}
\maketitle

\section{Definitions and theorems}
\subsection{Linear and integer programs}
The following is a \textbf{linear program}:
\begin{alignat*}{2}
    & \text{maximise} & \quad & c^Tx \\
    & \text{subject to} & & Ax \leq b
\end{alignat*}
Here $x, b, c\in \R^n$ and $A$ is an $m\times n$ matrix over $\R$. The function $c^Tx$ is called the \textbf{objective function} and it is what the linear program is trying to maximise (or minimise). Any vector $x$ that satisfies the inequality $Ax \leq b$ is called a \textbf{feasible solution}. A vector $x^*$ is called an \textbf{optimum solution} if $c^Tx^* \geq c^Tx$ for all feasible solutions $x$. A linear program is called \textbf{infeasible} if there are no feasible solutions. Even if a linear program is feasible, it might be \textbf{unbounded}. This happens when its objective function can be made arbitrarily "good".\\
\indent In a certain sense, linear programs are easy to solve. Adding an integrality constraint to a linear program gives an \textbf{integer program}:
\begin{alignat*}{2}
    & \text{maximise} & \quad & c^Tx \\
    & \text{subject to} & & Ax \leq b\\
    &                   & & x \in \Z^n
\end{alignat*}
Integer programs are much harder to solve in the general case. Suppose we want to solve the following integer program:
\begin{alignat*}{2}
    & \text{maximise} & \quad & c^Tx \\
    & \text{subject to} & & Ax \leq b\\
    &                   & & x \in \{0,1\}
\end{alignat*}
It is often a good strategy to solve the \textbf{linear program relaxation} of the integer program, which gives an approxmiate answer:
\begin{alignat*}{2}
    & \text{maximise} & \quad & c^Tx \\
    & \text{subject to} & & Ax \leq b\\
    &                   & & 0\leq x\leq 1
\end{alignat*}
Then we can take the solution of the linear program and round each element to the nearest integer value to get a solution of the original integer program. In some cases, the linear program relaxation gives quite a good answer. For example, if $G$ is a graph with minimum vertex cover of size $n$, then formulating an integer program and solving its linear program relaxation will spit out a vertex cover $S$ with $|S|\leq 2n$.

\subsection{Geometric interpretation}
The inequality $Ax\leq b$ defines a \textbf{polyhedron}. More precisely, we say that a polyhedron is the intersection of finitely many \textbf{halfspaces}, where a halfspace is the set $\{x : a^Tx\leq b\}$ for some $a\in \R^n$ and $b\in \R$. If a polyhedron is bounded, we call it a \textbf{polytope}. A polyhedron is closed and convex. The finite intersection of closed sets is closed and the intersection of convex sets is convex.\\
\indent Let $S\subset \R^n$ be a a set of $m$ points. Then the \textbf{convex hull} of $S$ is the set
$$\conv S = \{x : x = \sum_{i=1}^m\lambda_i s_i, \sum_{i=1}^m\lambda_i = 1, \lambda_i \geq 0\}.$$
The \textbf{conic hull} of $S$ is the set
$$\cone S = \{x : x = \sum_{i=1}^m\lambda_i s_i, \lambda_i\geq 0\}.$$
The \textbf{affine hull} of $S$ is the set
$$\aff S = \{x : x = \sum_{i=1}^m\lambda_i s_i, \sum_{i=1}^m\lambda_i = 1\}.$$
The convex hull can be obtained from the conic hull by intersecting it with a hyperplane. The affine hull is the smallest-dimensional subspace that contains all the points in $S$.\\
\indent Let $P = \{x: Ax\leq b\}$ be a polyhedron. An inequality $\alpha^Tx \leq \beta$ is said to be a \textbf{valid inequality} if and only if for all $y\in P$, $\alpha^T y\leq \beta$. Then a subset $F\subset P$ is a \textbf{face} of $P$ if and only if there exists a valid inequality $\alpha^T x\leq \beta$ such that $F = P\cap \{x : \alpha^T x\leq \beta\}$. The trivial faces of any polyhedron $P$ are $\emptyset$ and $P$ itself (formed by intersecting $P$ with the valid inequalities $0^Tx \leq 1$ and $0^Tx\leq 0$ respectively). Any face that is not trivial is called \textbf{proper}.\\
\indent The \textbf{dimension} of a polyhedron $P$, denoted $\dim P$, is the largest dimension of a ball inside $P$. A \textbf{vertex} is a face of dimension 0, an \textbf{edge} is a face of dimension 1, and a \textbf{facet} is a face of dimension $\dim P - 1$. A point $x\in P$ is called an \textbf{extreme point} if it is not a convex combination of points of $P$. Note that a point $x$ is an extreme point if and only if it is a vertex.
\begin{theorem*}{\textup{(Minkowski-Weyl)}}
    A set of points $P$ is a polyhedron if and only if there exist finite sets $V$ and $Y$ such that $P = \conv V + \cone Y$ where the sum $A + B$ denotes the set $\{a + b : a\in A, b\in B\}$.
\end{theorem*}

\subsection{Simplex method}
A polyhedron is \textbf{pointed} if it does not contain a line.
\begin{lemma*}
    Every non-empty pointed polyhedron $P$ has at least one vertex.
\end{lemma*}
\begin{theorem*}
    Consider the linear program
    \begin{alignat*}{2}
        & \textup{maximise} & \quad & c^Tx \\
        & \textup{subject to} & & Ax \leq b.
    \end{alignat*}
    If the polyhedron $P$ defined by $Ax\leq b$ is pointed and if an optimum exists, then there exists a vertex of $P$ that attains the maximum.
\end{theorem*}
The result of the above statements is an algorithm for solving linear programs, called the \textbf{simplex method}. Here is the general idea:
\begin{enumerate}
    \item Find an initial vertex. It is a 0-dimensional face with some 1-dimensional faces incident on it.
    \item At any vertex, check for optimality. If optimal, we are done.
    \item If the vertex is not optimal, increase the value of the objective function by following an edge to a new vertex. Return to step 2.
\end{enumerate}
\indent Let $A$ be an $m\times n$ matrix, $b\in \R^m$, $x \in\R^n$, with $n \geq m$. Let $B\subset [n] =\{1,\ldots,n\}$ of cardinality $m$. Then $A_B$ denotes the square matrix formed by selecting the columns of $A$ indexed by the elements of $B$ and likewise $x_B$ denotes the vector of length $m$ formed by selecting the elements indexed by the elements of $B$. We define a vector $x$ to be a \textbf{basic feasible solution} of the linear program
\begin{alignat*}{2}
    & \textup{maximise} & \quad & c^Tx \\
    & \textup{subject to} & & Ax \leq b
\end{alignat*}
if and only if there exists $B\subset [n]$ with $|B|= m$ such that:
\begin{enumerate}[i)]
    \item $A_B$ is non-singular
    \item $x_j = 0$ for all $j\notin B$
    \item $x_B = A_B^{-1}b$
\end{enumerate}
A set $B\subset [n]$ is called a \textbf{basis} if and only if $A_B$ is non-singular, and $B$ is a \textbf{feasible basis} if the unique solution of $A_Bx_B=b$ is non-negative. Note that a basic feasible solution is uniquely determined by the feasible basis $B$, but a vertex does \textit{not} have to be determined by a unique basis.\\
\indent A matrix $A$ is \textbf{totally unimodular} if and only if every square submatrix of $A$ has determinant 0, 1, or $-1$.

\subsection{Duality}
As always, we start with the linear program
\begin{alignat*}{2}
    & \textup{maximise} & \quad & c^Tx \\
    & \textup{subject to} & & Ax \leq b\\
    &                     & & x \geq 0.
\end{alignat*}
Inequalities of a linear program can be added together without changing the feasible region. Likewise, one can also multiply any inequality by a non-negative scalar. So we can manipulate inequalities to get an upper bound on the optimum value of the objective function. We want to pick coefficients to get the \textit{least} upper bound possible. Let $y$ denote a vector of coefficients. We want to
\begin{alignat*}{2}
    & \textup{minimise} & \quad & b^Ty \\
    & \textup{subject to} & & A^Ty \geq c\\
    &                     & & y \geq 0
\end{alignat*}
The original linear program is called the \textbf{primal} and this new one is called the \textbf{dual}. Any feasible solution of the dual linear program gives an upper bound on the primal and the optimum of the dual is equal to the optimum of the original program.

\begin{theorem*}{\textup{(Weak Duality Theorem)}}
    Let $P$ and $D$ be primal and dual linear programs of the above form and let $x$ and $y$ be respective feasible solutions. Then $c^Tx \leq b^Ty$.
\end{theorem*}

\begin{theorem*}{\textup{(Strong Duality Theorem)}}
    Let $P$ and $D$ be primal and dual linear programs. Then exactly one of the following holds:
    \begin{enumerate}[i)]
        \item P infeasible, D infeasible
        \item P infeasible, D unbounded
        \item P unbounded, D infeasible
        \item P and D both feasible and bounded and furthermore, if $x^*$ and $y^*$ are primal and dual optimal solutions, then $c^Tx^* = b^Ty^*$.
    \end{enumerate}
\end{theorem*}

\begin{lemma*}{\textup{(Farkas lemma)}}
    For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
    \begin{enumerate}[i)]
        \item There exists a vector $x$ such that $Ax = b$ and $x \geq 0$.
        \item There exists a vector $y$ such that $A^Ty \geq 0$ and $b^Ty< 0$.
    \end{enumerate}
\end{lemma*}

\begin{theorem*}{\textup{(Farkas lemma 2.0)}}
    For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
    \begin{enumerate}[i)]
        \item There exists a vector $x$ such that $Ax \leq b$ and $x \geq 0$.
        \item There exists a vector $y$ such that $A^Ty \geq 0$, $y\geq0$ and $b^Ty< 0$.
    \end{enumerate}
\end{theorem*}

\subsection{Applications}
\subsubsection{Matching and vertex cover in bipartite graphs}
Given a graph $G=(V,E)$, a \textbf{matching} is a set $M\subset E$ such that for all $v\in V$ there is at most one edge $e\in M$ with $v\in e$. The linear program that finds the maximal matching in a bipartite graph is
\begin{alignat*}{2}
    & \textup{maximise} & \quad & \sum_{e\in E} x_e \\
    & \textup{subject to} & & \sum_{v\in e}x_e \leq 1,\ \text{for all}\ v\in V\\
    &                     & & x_e \geq 0,\ \text{for all}\ e\in E.
\end{alignat*}
The dual of this linear program is the following:
\begin{alignat*}{3}
    & \textup{minimise} & \quad & \sum_{v\in V} y_v &\\
    & \textup{subject to} & & y_u + y_v \geq 1,\  \text{for all}\ \{u,v\}\in E\\
    &                     & & y_v \geq 0, \  \text{for all}\ v\in V.
\end{alignat*}
    Given a graph $G=(V,E)$, a \textbf{vertex cover} is a set $S \subset V$ such that for all $\{u, v\}\in E$, $\{u,v\}\cap S\geq 1$. The dual linear program above finds the minimum vertex cover in a bipartite graph. The duality of these two programs results in K\H{o}nig's theorem.
\begin{theorem*}{\textup{(K\H{o}nig's theorem)}}
    Suppose $G=(V,E)$ is a bipartite graph. Let $M\subset E$ denote the maximum matching and let $S\subset V$ denote the minimum vertex cover. Then $|M|=|S|$.
\end{theorem*}

\subsubsection{Matching in weighted graphs}
Let $G=(V,E)$ be a graph (not necessarily bipartite) with a weight function $w:E\rightarrow \R^+$. Suppose we want to find the maximum weight matching in $G$. For any subset $E'\subset E$, the \textbf{characteristic vector} of $E'$ is the vector $\chi^{E'}$ given by
$$ \chi_e^{E'} = \begin{cases} 1 & e\in E' \\ 0 & e\notin E'\end{cases} $$
for all $e\in E$. Then the \textbf{matching polytope} of the graph $G$ is the set
$$ M(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \text{is the characteristic vector of some matching in}\ G\}.$$
To get the maximum weight matching, we need only solve the linear program
\begin{alignat*}{2}
    & \textup{maximise} & \quad & w^Tx \\
    & \textup{subject to} & & x\in M(G).
\end{alignat*}
What does the polytope $M(G)$ look like in terms of inequalities? It turns out that $M(G)$ is the set of all vectors $x$ that satisfy
\begin{enumerate}
    \item $\displaystyle\sum_{v\in e}x_e \leq 1$ for all $v\in V$ and
    \item $\displaystyle\sum_{e\in E[U]}x_e \leq \frac{|U|-1}{2}$ for every set $U\subset V$ of odd cardinality.
\end{enumerate}
In a graph $G=(V,E)$, a matching $M\subset E$ is \textbf{perfect} if every vertex in $V$ is incident to some edge in $M$. The perfect matching polytope is the set
$$ PM(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \text{is the characteristic vector of some perfect matching in}\ G\}.$$
For a set $U\subset V$ of vertices, let $\delta(U)$ denote the number of edges that have one endpoint in $U$ and one endpoint outside $U$. Then we find that $PM(G)$ is the set of all vectors $x$ satisfying
\begin{enumerate}
    \item $\displaystyle\sum_{v\in e}x_e = 1$ for all $v\in V$,
    \item $\displaystyle\sum_{e\in \delta(U)}x_e\geq 1$ for every set $U\subset V$ of odd cardinality, and
    \item $x_e \geq 0$ for every $e\in E$.
\end{enumerate}
Note that the number of inequalities in these polytopes may be very large. For this reason, a polynomial-time algorithm may not exist.

\subsubsection{The travelling salesman problem}
Given a graph $G=(V,E)$, a \textbf{closed tour} is a walk in $G$ with no repeated edges and no repeated vertices except for the first and last vertex, which are the same. In a graph with edge costs $(c_e : e\in E)$, \textbf{travelling salesman problem} asks us to find the minimum-weight closed tour that contains each vertex exactly once. To solve the travelling salesman problem, we start with the following linear program:
\begin{alignat*}{2}
    & \textup{minimise} & \quad & \sum_{e\in E} c_ex_e \\
    & \textup{subject to} & & \sum_{v\in e} = 2\ \text{for all}\ v\in V,\\
    &  & & 0\leq x_e \leq 1\ \text{for all}\ e\in E.
\end{alignat*}
What we really want to do is add the inequalities $\displaystyle\sum_{e\in \delta(S)} x_e \geq 2$ for every $\emptyset \neq S \neq V$. This is called \textbf{subtour elimination} because it forbids any tours smaller than the entire closed tour we're trying to build. Adding this inequality, we get the linear program relaxation of the integer program for the travelling salesman problem. The problem is that eliminating subtours adds an exponential number of ineqalities to the program.\\
\indent Instead of adding all of these inequalities, we use the cutting-plane approach. Given a polytope $P$, a \textbf{cutting plane} is an inequality that is valid for all integer points in $P$. It is called a cutting plane because it may cut the original polytope. We can apply this idea as follows:
\begin{enumerate}
    \item We start with the linear program without the subtour constraints.
    \item Find the optimal solution to the linear program.
    \item If we find that some subtour constraints are violated by the optimal solution, we add these cutting planes to our program and return to step 2. If no subtour constraints are violated, we have a solution and we may stop.
\end{enumerate}
Step 3 is our main concern. How do we figure out if there are any violated subtour constraints? Suppose we are at a certain step of the algorithm and the linear program has just given us an optimal solution $x^*$. Let $G(x^*)$ denote the graph with vertices $V$ and edges $\{e : x_e^* > 0\}$. Without loss of generality, assume $G(x^*)$. Set capacities $u_e = x_e^*$ for every edge $e$. Now for any $S\subset V$, the weight of the cut $(S, V\setminus S)$ equals the value of $\displaystyle\sum_{e\in \delta(S)} x_e$. So we can solve the min-cut problem over $G(x^*)$ and if we get a value less than 2, we know that some subtour constraint is violated. Of course, we need some heuristics to decide which constraints to add, so there is still no guarantee of polynomial-time termination.

>>>>

% Notes for NOPT048 Linear Programming and Combinatorial Optimization
% Charles University in Prague, Summer 2019
% Instructor: Prof. Hans Raj Tiwary
% Notes by Marcel Goh

\input fontmac
\input mathmac

\def\conv{\hbox{conv}\,}
\def\cone{\hbox{cone}\,}
\def\aff{\hbox{aff}\,}

\def\maximise{\hbox{maximise}\hfill}
\def\minimise{\hbox{minimise}\hfill}
\def\subject{\hbox{subject to}\hfill}
\def\hcr{\hfill\cr}

\leftrighttop{NOPT048 LINEAR PROGRAMMING AND COMBINATORIAL OPTIMIZATION}{MARCEL GOH}

\maketitle{NOPT048 Linear Programming and Combinatorial Optimization}{Notes by}{Marcel K. Goh {\rm (Prague, Czech Rep.)}}{9 June 2019}

\section 1. DEFINITIONS AND THEOREMS

\vskip -\medskipamount

\section 1.1. Linear and integer programs

The following is a {\it linear program}:
$$\matrix{
    & \maximise & \quad & c^Tx\hcr
    & \subject & & Ax \leq b.\hfill
}$$

Here $x, b, c\in \R^n$ and $A$ is an $m\times n$ matrix over $\R$. The function $c^Tx$ is called the {\it objective function} and it is what the linear program is trying to maximise (or minimise). Any vector $x$ that satisfies the inequality $Ax \leq b$ is called a {\it feasible solution}. A vector $x^*$ is called an {\it optimum solution} if $c^Tx^* \geq c^Tx$ for all feasible solutions $x$. A linear program is called {\it infeasible} if there are no feasible solutions. Even if a linear program is feasible, it might be {\it unbounded}. This happens when its objective function can be made arbitrarily ``good''.

In a certain sense, linear programs are easy to solve. Adding an integrality constraint to a linear program gives an {\it integer program}:
$$\matrix{
    & \maximise & \quad & c^Tx \hcr
    & \subject & & Ax \leq b,\hcr
    &                   & & x \in \Z^n.\hfill
}$$
Integer programs are much harder to solve in the general case. Suppose we want to solve the following integer program:
$$\matrix{
    & \maximise & \quad & c^Tx \hcr
    & \subject & & Ax \leq b,\hcr
    &                   & & x \in \{0,1\}.\hfill
}$$
It is often a good strategy to solve the {\it linear program relaxation} of the integer program, which gives an approxmiate answer:
$$\matrix{
    & \maximise & \quad & c^Tx \hcr
    & \subject & & Ax \leq b,\hcr
    &                   & & 0\leq x\leq 1.\hfill
}$$
Then we can take the solution of the linear program and round each element to the nearest integer value to get a solution of the original integer program. In some cases, the linear program relaxation gives quite a good answer. For example, if $G$ is a graph with minimum vertex cover of size $n$, then formulating an integer program and solving its linear program relaxation will spit out a vertex cover $S$ with $|S|\leq 2n$.

\section 1.2. Geometric interpretation

The inequality $Ax\leq b$ defines a {\it polyhedron}. More precisely, we say that a polyhedron is the intersection of finitely many {\it halfspaces}, where a halfspace is the set $\{x : a^Tx\leq b\}$ for some $a\in \R^n$ and $b\in \R$. If a polyhedron is bounded, we call it a {\it polytope}. A polyhedron is closed and convex. The finite intersection of closed sets is closed and the intersection of convex sets is convex.

Let $S\subset \R^n$ be a a set of $m$ points. Then the {\it convex hull} of $S$ is the set
$$\conv S = \{x : x = \displaystyle\sum_{i=1}^m\lambda_i s_i, \displaystyle\sum_{i=1}^m\lambda_i = 1, \lambda_i \geq 0\}.$$
The {\it conic hull} of $S$ is the set
$$\cone S = \{x : x = \displaystyle\sum_{i=1}^m\lambda_i s_i, \lambda_i\geq 0\}.$$
The {\it affine hull} of $S$ is the set
$$\aff S = \{x : x = \displaystyle\sum_{i=1}^m\lambda_i s_i, \displaystyle\sum_{i=1}^m\lambda_i = 1\}.$$
The convex hull can be obtained from the conic hull by intersecting it with a hyperplane. The affine hull is the smallest-dimensional subspace that contains all the points in $S$.

Let $P = \{x: Ax\leq b\}$ be a polyhedron. An inequality $\alpha^Tx \leq \beta$ is said to be a {\it valid inequality} if and only if for all $y\in P$, $\alpha^T y\leq \beta$. Then a subset $F\subset P$ is a {\it face} of $P$ if and only if there exists a valid inequality $\alpha^T x\leq \beta$ such that $F = P\cap \{x : \alpha^T x\leq \beta\}$. The trivial faces of any polyhedron $P$ are $\emptyset$ and $P$ itself (formed by intersecting $P$ with the valid inequalities $0^Tx \leq 1$ and $0^Tx\leq 0$ respectively). Any face that is not trivial is called {\it proper}.

The {\it dimension} of a polyhedron $P$, denoted $\dim P$, is the largest dimension of a ball inside $P$. A {\it vertex} is a face of dimension 0, an {\it edge} is a face of dimension 1, and a {\it facet} is a face of dimension $\dim P - 1$. A point $x\in P$ is called an {\it extreme point} if it is not a convex combination of points of $P$. Note that a point $x$ is an extreme point if and only if it is a vertex.

\thmparens{M}{Minkowski-Weyl} A set of points $P$ is a polyhedron if and only if there exist finite sets $V$ and $Y$ such that $P = \conv V + \cone Y$ where the sum $A + B$ denotes the set $\{a + b : a\in A, b\in B\}$.

\section 1.3. Simplex method

A polyhedron is {\it pointed} if it does not contain a line.
\proclaim Lemma. Every non-empty pointed polyhedron $P$ has at least one vertex.

\proclaim Theorem. Consider the linear program
$$\matrix{
    & \hbox{\rm maximise}\hfill & \quad & c^Tx \hcr
    & \hbox{\rm subject to}\hfill & & Ax \leq b.\hfill
}$$
If the polyhedron $P$ defined by $Ax\leq b$ is pointed and if an optimum exists, then there exists a vertex of $P$ that attains the maximum.

The result of the above statements is an algorithm for solving linear programs, called the {\it simplex method}. Here is the general idea:

\algbegin Algorithm X (Simplex method). Find an optimum solution of a linear program.

\algstep X1. Find an initial vertex. It is a 0-dimensional face with some 1-dimensional faces incident on it.
\algstep X2. At any vertex, check for optimality. If optimal, we are done.
\algstep X3. If the vertex is not optimal, increase the value of the objective function by following an edge to a new vertex. Return to step X2.\slug

Let $A$ be an $m\times n$ matrix, $b\in \R^m$, $x \in\R^n$, with $n \geq m$. Let $B\subset [n] =\{1,\ldots,n\}$ of cardinality $m$. Then $A_B$ denotes the square matrix formed by selecting the columns of $A$ indexed by the elements of $B$ and likewise $x_B$ denotes the vector of length $m$ formed by selecting the elements indexed by the elements of $B$. We define a vector $x$ to be a {\it basic feasible solution} of the linear program
$$\matrix{
    & \maximise & \quad & c^Tx \hcr
    & \subject & & Ax \leq b\hfill
}$$
if and only if there exists $B\subset [n]$ with $|B|= m$ such that:
\medskip
\item {i)} $A_B$ is non-singular;
\smallskip
\item {ii)} $x_j = 0$ for all $j\notin B$;
\smallskip
\item {iii)} $x_B = A_B^{-1}b$.
\medskip
A set $B\subset [n]$ is called a {\it basis} if and only if $A_B$ is non-singular, and $B$ is a {\it feasible basis} if the unique solution of $A_Bx_B=b$ is non-negative. Note that a basic feasible solution is uniquely determined by the feasible basis $B$, but a vertex does {\it not} have to be determined by a unique basis.

A matrix $A$ is {\it totally unimodular} if and only if every square submatrix of $A$ has determinant 0, 1, or $-1$.

\section 1.4. Duality

As always, we start with the linear program
$$\matrix{
    & \maximise & \quad & c^Tx \hcr
    & \subject & & Ax \leq b,\hcr
    &                     & & x \geq 0.\hfill\hfill
}$$
Inequalities of a linear program can be added together without changing the feasible region. Likewise, one can also multiply any inequality by a non-negative scalar. So we can manipulate inequalities to get an upper bound on the optimum value of the objective function. We want to pick coefficients to get the {\it least} upper bound possible. Let $y$ denote a vector of coefficients. We want to
$$\matrix{
    & \minimise & \quad & b^Ty \hcr
    & \subject & & A^Ty \geq c,\hcr
    &                     & & y \geq 0.\hfill
}$$
The original linear program is called the {\it primal} and this new one is called the {\it dual}. Any feasible solution of the dual linear program gives an upper bound on the primal and the optimum of the dual is equal to the optimum of the original program.

\thmparens{W}{Weak Duality Theorem} Let $P$ and $D$ be primal and dual linear programs of the above form and let $x$ and $y$ be respective feasible solutions. Then $c^Tx \leq b^Ty$.

\thmparens{S}{Strong Duality Theorem} Let $P$ and $D$ be primal and dual linear programs. Then exactly one of the following holds:
\medskip
\item {i)} P infeasible, D infeasible
\smallskip
\item {ii)} P infeasible, D unbounded
\smallskip
\item {iii)} P unbounded, D infeasible
\smallskip
\item {iv)} P and D both feasible and bounded and furthermore, if $x^*$ and $y^*$ are primal and dual optimal solutions, then $c^Tx^* = b^Ty^*$.
\medskip

\lemparens{F}{Farkas lemma}
For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
\medskip
\item {i)} There exists a vector $x$ such that $Ax = b$ and $x \geq 0$.
\smallskip
\item {ii)} There exists a vector $y$ such that $A^Ty \geq 0$ and $b^Ty< 0$.
\medskip

\lemparens{G}{Farkas lemma 2.0}
For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
\medskip
\item {i)} There exists a vector $x$ such that $Ax \leq b$ and $x \geq 0$.
\smallskip
\item {ii)} There exists a vector $y$ such that $A^Ty \geq 0$, $y\geq0$ and $b^Ty< 0$.
\medskip

\section 1.5. Applications

\vskip -\medskipamount

\section 1.5.1. Matching and vertex cover in bipartite graphs

Given a graph $G=(V,E)$, a {\it matching} is a set $M\subset E$ such that for all $v\in V$ there is at most one edge $e\in M$ with $v\in e$. The linear program that finds the maximal matching in a bipartite graph is
$$\matrix{
    & \maximise   & \quad & \displaystyle\sum_{e\in E} x_e  \hfill    & \hcr
    & \subject &       & \displaystyle\sum_{v\in e}x_e \leq 1  \hfill &\hbox{for all}\ v\in V,\hcr
    &                   &       & x_e \geq 0      \hfill          &\hbox{for all}\ e\in E.\hfill
}$$
The dual of this linear program is the following:
$$\matrix{
    & \minimise   & \quad & \displaystyle\sum_{v\in V} y_v  \hfill & \hcr
    & \subject &       & y_u + y_v \geq 1   \hfill &\hbox{for all}\ \{u,v\}\in E,\hcr
    &                   &       & y_v \geq 0        \hfill  &\hbox{for all}\ v\in V.\hfill
}$$
    Given a graph $G=(V,E)$, a {\it vertex cover} is a set $S \subset V$ such that for all $\{u, v\}\in E$, $\{u,v\}\cap S\geq 1$. The dual linear program above finds the minimum vertex cover in a bipartite graph. The duality of these two programs results in K\H{o}nig's theorem.
\thmparens{K}{K\H{o}nig's theorem}
    Suppose $G=(V,E)$ is a bipartite graph. Let $M\subset E$ denote the maximum matching and let $S\subset V$ denote the minimum vertex cover. Then $|M|=|S|$.\slug

\section 1.5.2. Matching in weighted graphs

Let $G=(V,E)$ be a graph (not necessarily bipartite) with a weight function $w:E\rightarrow \R^+$. Suppose we want to find the maximum weight matching in $G$. For any subset $E'\subset E$, the {\it characteristic vector} of $E'$ is the vector $\chi^{E'}$ given by
$$\chi_e^{E'} = \cases{
    1 & if $e\in E'$ \hcr
    0 & if $e\notin E'$\hfill
}$$
for all $e\in E$. Then the {\it matching polytope} of the graph $G$ is the set
$$ M(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \hbox{is the characteristic vector of some matching in}\ G\}.$$
To get the maximum weight matching, we need only solve the linear program
$$\matrix{
    & \maximise & \quad & w^Tx \hcr
    & \subject & & x\in M(G).\hfill
}$$
What does the polytope $M(G)$ look like in terms of inequalities? It turns out that $M(G)$ is the set of all vectors $x$ that satisfy
\medskip
\item {1.} $\displaystyle\sum_{v\in e}x_e \leq 1$ for all $v\in V$ and
\smallskip
\item {2.} $\displaystyle\sum_{e\in E[U]}x_e \leq {|U|-1\over 2}$ for every set $U\subset V$ of odd cardinality.
\medskip
In a graph $G=(V,E)$, a matching $M\subset E$ is {\it perfect} if every vertex in $V$ is incident to some edge in $M$. The perfect matching polytope is the set
$$ PM(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \hbox{is the characteristic vector of some perfect matching in}\ G\}.$$
For a set $U\subset V$ of vertices, let $\delta(U)$ denote the number of edges that have one endpoint in $U$ and one endpoint outside $U$. Then we find that $PM(G)$ is the set of all vectors $x$ satisfying
\medskip
\item {1.} $\displaystyle\sum_{v\in e}x_e = 1$ for all $v\in V$,
\smallskip
\item {2.} $\displaystyle\sum_{e\in \delta(U)}x_e\geq 1$ for every set $U\subset V$ of odd cardinality, and
\smallskip
\item {3.} $x_e \geq 0$ for every $e\in E$.
\medskip
Note that the number of inequalities in these polytopes may be very large. For this reason, a polynomial-time algorithm may not exist.

\section 1.5.3. The travelling salesman problem

Given a graph $G=(V,E)$, a {\it closed tour} is a walk in $G$ with no repeated edges and no repeated vertices except for the first and last vertex, which are the same. In a graph with edge costs $(c_e : e\in E)$, {\it travelling salesman problem} asks us to find the minimum-weight closed tour that contains each vertex exactly once. To solve the travelling salesman problem, we start with the following linear program:
$$\matrix{
    & \minimise & \quad & \displaystyle\sum_{e\in E} c_ex_e \hcr
    & \subject & & \displaystyle\sum_{v\in e} = 2\ \hbox{for all}\ v\in V,\hcr
    &  & & 0\leq x_e \leq 1\ \hbox{for all}\ e\in E.\hfill
}$$
What we really want to do is add the inequalities $\sum_{e\in \delta(S)} x_e \geq 2$ for every $\emptyset \neq S \neq V$. This is called {\it subtour elimination} because it forbids any tours smaller than the entire closed tour we're trying to build. Adding this inequality, we get the linear program relaxation of the integer program for the travelling salesman problem. The problem is that eliminating subtours adds an exponential number of ineqalities to the program.
Instead of adding all of these inequalities, we approach the problem as follows:

\algbegin Algorithm T (Subtour elimination). This algorithm solves the travelling salesman problem by adding violated subtour constraints one at a time.

\algstep T1. We start with the linear program without the subtour constraints.
\algstep T2. Find the optimal solution to the linear program.
\algstep T3. If we find that some subtour constraints are violated by the optimal solution, we add these inequalities to our program and return to step T2. If no subtour constraints are violated, we have a solution.\slug

Step 3 is our main concern. How do we figure out if there are any violated subtour constraints? Suppose we are at a certain step of the algorithm and the linear program has just given us an optimal solution $x^*$. Let $G(x^*)$ denote the graph with vertices $V$ and edges $\{e : x_e^* > 0\}$. Without loss of generality, assume $G(x^*)$. Set capacities $u_e = x_e^*$ for every edge $e$. Now for any $S\subset V$, the weight of the cut $(S, V\setminus S)$ equals the value of $\sum_{e\in \delta(S)} x_e$. So we can solve the min-cut problem over $G(x^*)$ and if we get a value less than 2, we know that some subtour constraint is violated. Of course, we need some heuristics to decide which constraints to add, so there is still no guarantee of polynomial-time termination.

\section 1.6. Ellipsoid algorithm

So far, we do not know if linear programs can be solved in polynomial time. It turns out that optimisation is polynomially equivalent to separating over a polytope. Given $x\in \R^n$ and a polyhedron $P$, we separate as follows: Output ``yes'' if $x\in P$; otherwise, output ``no'' and a pair $(\alpha, \beta)$ such that $\alpha^Ty\leq\beta$ is a valid inequality for $P$ and $\alpha^T x > \beta$. The {\it ellipsoid algorithm} uses an oracle for separation to answer the question of feasibility: Given a polytope $P$, find an $x\in P$ or conclude that $P = \emptyset$.

Before we get into the actual method, let's convince ourselves that checking for feasibility is equivalent to optimising. That optimisation implies feasibility is obvious. So let's suppose we have a method to check if a linear program is feasible. Can we use that to optimise the linear program
$$\matrix{
    & \minimise & \quad & c^Tx\hcr
    & \subject & & Ax \leq b,\hcr
    &  & & x\geq 0?\hfill
}$$
Well, from duality we know that if an optimum exists, it will also be the optimum value of the dual linear program. So we can construct a polyhedron that contains only this value and then check if it is feasible. So we check if the polyhedron
$$\{\pmatrix{ x\hcr y } : c^Tx = b^Ty, Ax \leq b, A^Ty \geq c, x \geq 0, y\geq 0\}$$
is feasible and that will give us the optimum value of $x$, if it exists.

On to ellipsoids. An {\it ellipsoid} $E(A,a)$ is the set $\{x \in \R^n : (x - a)^TA^{-1}(x-a) \leq 1\}$ where $A$ is a {\it positive definite} matrix. This means that $(x-a)^TA^{-1}(x-a)$ is positive and moreover, $A = M\cdot M$ for some square matrix $M$. If we let $B$ denote the unit ball in $\R^n$, then $E(A,a) = \{Mx + a : x \in B\}$.

For simplicity, assume that $P$ is bounded and assume that $P$ is either empty or full-dimensional. Assume also that we have an oracle that outputs a separating hyperplane. Then the {\it ellipsoid algorithm} determines if $P = \{x : Ax \leq b\}$ is feasible.
